---
title: "Reproducible Research Course Project 1"
author: "Victor Paul dela Cruz"
date: "December 19, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Loading Processing Data

##1. Show Any code that is needed of the following:
###1.1 Load the data (i.e. read.csv()) 
###1.2Process the transform the data into a formal suitable for the analyses
```{r, include=TRUE}

setwd("D:/victorstat/MAJORS/Reproducible Research")
# Import data
activity <- read.csv("activity.csv")
# libraries
library(ggplot2)
library(dplyr)
Sys.setlocale("LC_TIME", "English")

#Description of the Variable
str(activity)

```

##As we observe, the variables included in this dataset are:
###1. steps: Number of steps taking in a 5-minute interval (missing values are coded as NA)
###2. date: The date on which the measurement was taken in YYYY-MM-DD format
###3. interval: Identifier for the 5-minute interval in which measurement was taken

##2.Total Number of steps taken per day 
###2.1 Calculate the total number of steps
###2.2 Make a histogram of the total number of step taken each day
###2.3 Calculate and report the mean and median total number of steps per day

```{r numbersteps}
#Number 2.1
StepsPerDay <- aggregate(activity$steps, list(activity$date), FUN=sum)
colnames(StepsPerDay) <- c("Date", "Steps")
StepsPerDay

#Number 2.2
# draw the histogram
g <- ggplot(StepsPerDay, aes(Steps))
g+geom_histogram(boundary=0, binwidth=2500, col="darkgreen", fill="lightgreen")+ggtitle("Histogram of steps per day")+xlab("Steps")+ylab("Frequency")+theme(plot.title = element_text(face="bold", size=12))+scale_x_continuous(breaks=
                                                                                                                                seq(0,25000,2500))+scale_y_continuous(breaks=seq(0,18,2))
#Mean and Median
mean(StepsPerDay$Steps, na.rm=TRUE)
median(StepsPerDay$Steps, na.rm=TRUE)
```

##3. Average daily activity pattern
###3.1 Make time series plot of the five-minute interval and average number of steps taken
###3.2 Which five-minute interval, on average across all days in the dataset, contains the maximum number of steps
```{r}
# create table with steps per time
StepsPerTime <- aggregate(steps~interval,data=activity,FUN=mean,na.action=na.omit)
# variable time (more comprensible for the graph axis)
StepsPerTime$time <- StepsPerTime$interval/100
# draw the line plot
h <- ggplot(StepsPerTime, aes(time, steps))
h+geom_line(col="brown")+ggtitle("Average steps per time interval")+xlab("Time")+ylab("Steps")+theme(plot.title = element_text(face="bold", size=12))

# table for dplyr
ST <- tbl_df(StepsPerTime)
# find the column
ST %>% select(time, steps) %>% filter(steps==max(ST$steps))

```
##4. Imputing missing values
###Note that there are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.
###1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
###2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.
###3. Create a new dataset that is equal to the original dataset but with the missing data filled in.
###4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?
```{r}
#for dplyr
ACT <- tbl_df(activity)
# find the column
ACT %>% filter(is.na(steps)) %>% summarize(missing_values = n())

```
###Rounded Values is used for the missing observations
```{r}
# values without NA are imputed in a new column
activity$CompleteSteps <- ifelse(is.na(activity$steps), round(StepsPerTime$steps[match(activity$interval, StepsPerTime$interval)],0), activity$steps)

activityFull <- data.frame(steps=activity$CompleteSteps, interval=activity$interval, date=activity$date)
# see first 10 values of the new dataset
head(activityFull, n=10)
```
###Histogram of the total number of steps per day with missing values filled in
```{r}
# prepare data
StepsPerDayFull <- aggregate(activityFull$steps, list(activityFull$date), FUN=sum)
colnames(StepsPerDayFull) <- c("Date", "Steps")
# draw the histogram
g <- ggplot(StepsPerDayFull, aes(Steps))
g+geom_histogram(boundary=0, binwidth=2500, col="darkblue", fill="lightblue")+ggtitle("Histogram of steps per day")+xlab("Steps")+ylab("Frequency")+theme(plot.title = element_text(face="bold", size=12))+scale_x_continuous(breaks=seq(0,25000,2500))+scale_y_continuous(breaks=seq(0,26,2))
```
###Mean and Median of recent filled observations
```{r}
mean(StepsPerDayFull$Steps)
median(StepsPerDayFull$Steps)
```
###Imputing missing data have only a little and transcurable impact on the mean ant the median of the total daily number of steps. Watching the histogram we can note than the only bin that is changed is the interval between 10000 and 12500 steps, grown from a frequency of 18 to a frequency of 26. Different methods for replace missing values could cause different results.

##5. Are there differences in activity patterns between weekdays and weekends?
###For this part the weekdays() function may be of some help here. Use the dataset with the filled-in missing values for this part.
###1. Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.
###2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

###Creating a new factor variables weekdays and weekends
```{r}
# Create variable with date in correct format
activityFull$RealDate <- as.Date(activityFull$date, format = "%Y-%m-%d")
# create a variable with weekdays name
activityFull$weekday <- weekdays(activityFull$RealDate)
# create a new variable indicating weekday or weekend
activityFull$DayType <- ifelse(activityFull$weekday=='Saturday' | activityFull$weekday=='Sunday', 'weekend','weekday')
# see first 10 values
head(activityFull, n=10)
```

###Time Series Plot: Weekday vs. Weekend
```{r}
# create table with steps per time across weekdaydays or weekend days
StepsPerTimeDT <- aggregate(steps~interval+DayType,data=activityFull,FUN=mean,na.action=na.omit)
# variable time (more comprensible for the graph axis)
StepsPerTimeDT$time <- StepsPerTime$interval/100
# draw the line plot
j <- ggplot(StepsPerTimeDT, aes(time, steps))
j+geom_line(col="darkred")+ggtitle("Average steps per time interval: weekdays vs. weekends")+xlab("Time")+ylab("Steps")+theme(plot.title = element_text(face="bold", size=12))+facet_grid(DayType ~ .)
```